---
title: "Evaluation of a split season team ability model"
---

```{r setup, include = FALSE}
source("setup.R")

library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)

library(rstan)
library(bayesplot)
color_scheme_set(c("purple"))

source("plot-utils.R")
theme_setup()
```

I've now [extended] the previous [single-season, single ability model]
to estimate abilities at different splits in the season.  The new
model largely follows Milad Kharratzadeh's English Premier League
model, with "weeks" corresponding to "periods".  The most substantial
difference between the EPL model and the one here is that a team's
ability is not represented as a random walk where the ability from the
previous period serves as the mean for the next period's ability.
Instead, a team's ability values for all periods are modeled as
normally distributed around a common mean.

[extended]: scorediff-split.stan
[single-season, single ability model]: scorediff-oneseason-2011.html
[English Premier League model]: https://github.com/milkha/EPL/blob/master/epl_model.stan

I've run the model with the 2011 game logs.  I split the season into
eight periods (~20 games), thinking that it would do an OK job of
capturing the ups and downs a team goes through in a season while not
introducing an unwieldy number of parameters.  I haven't played around
with any other splits yet.

## Model diagnostics

```{r fit, cache.extra = file.info("../models/scorediff-split_2011-fit.rds")}
fit <- readRDS("../models/scorediff-split_2011-fit.rds")
```

Clicking around in ShinyStan UI (`launch_shinystan(fit)`), the
diagnostic statistics look fine overall.  Some parameters have
effective sample sizes on the lower end.

```{r neff_plot, dependson = "fit", out.width = "50%", fig.width = 4.28}
model_pars <- names(fit)
selpars <- model_pars[!grepl("^eta", model_pars) &
                      !grepl("score_diff_new", model_pars)]

nrats <- neff_ratio(fit, pars = selpars)
nrats[nrats < 0.5]

mcmc_neff_hist(nrats) + theme(panel.grid = element_blank())
```

## Parameter estimates

### t-distribution degrees of freedom

Following the EPL model, I've included a parameter for the degrees of
freedom used for the t-distribution.  Most of the estimated values are
larger than 7, which is the fixed value I used for in the previous
model.

```{r samps, dependson = "fit"}
samps <- rstan::extract(fit)
```

```{r nu_distr_plot, dependson = "samps", out.width = "50%", fig.width = 4.28}
quantile(samps$nu, probs = c(0.025, 0.5, 0.975))

data.frame(nu = samps$nu) %>%
    ggplot(aes(nu)) +
    geom_histogram(bins = 40) +
    scale_x_continuous(limits = c(5, NA)) +
    labs(x = expression(nu)) +
    theme_histogram()
```

With a median $\nu$ near 11, it probably makes sense to stick with a
t-distribution rather than switching to a normal.

```{r nu_normal_plot, fig.asp = 0.5}
set.seed(113724)
nvals <- tibble(x = seq(-6, 6, length.out = 500),
                y = dnorm(x))
tvals <- expand.grid(x = nvals$x, df = c(3, 7, 11, 17)) %>%
    mutate(y = dt(x, df = df),
           df = factor(df))

ggplot(tvals) +
    geom_line(aes(x = x, y = y, color = df)) +
    geom_line(data = mutate(nvals, distr = "normal"),
              aes(x = x, y = y, linetype = distr)) +
    labs(color = expression(nu)) +
    guides(color = guide_legend(order = 0),
           linetype = guide_legend(order = 1, title = NULL)) +
    theme_void() +
    theme(legend.position = c(0.2, 0.5)) +
    scale_linetype_manual(values = c("dashed")) +
    scale_color_manual(values = c("gray70", tc$primary_light,
                                  tc$primary_dim, tc$primary_dark))
```

Both $\nu$ and $\sigma_y$ influence the range of observed score
differentials, and, unsurprisingly, iterations with larger values for
$\nu$ tend to have larger values for $\sigma_y$.

```{r nu_sigma_y_plot, dependson = "samps", out.width = "50%", fig.width = 4.28}
data.frame(nu = samps$nu,
           sigma_y = samps$sigma_y) %>%
    ggplot(aes(nu, sigma_y)) +
    labs(x = expression(nu),
         y = expression(sigma[y])) +
    geom_point(alpha = 0.2, shape = 16) +
    theme_axis_line()
```

### Home-field advantage

The model includes a term for home-field avantage.  The estimates are
uncertain, small in magnitude, and slightly shifted in the direction
of the home team.  No suprises there.

```{r b_home_distr_plot, dependson = "samps", out.width = "50%", fig.width = 4.28}
quantile(samps$b_home, probs = c(0.025, 0.5, 0.975))

data.frame(b_home = samps$b_home) %>%
    ggplot(aes(b_home)) +
    geom_vline(xintercept = 0, size = 0.8, color = tc$primary_light_dim) +
    geom_histogram(bins = 40, size = 0.3) +
    labs(x = expression(beta[home])) +
    theme_histogram()
```

Perhaps this parameter should be estimated for each team in order to
catch park-specific effects.  That's a bit complicated because a
hitter's park can help both teams score more runs, but I suppose it
could capture the extent that the home team plays to (or fails to play
to) their park's strengths.

### Abilities

The main difference between this model and its predecessor is that a
single-season ability estimate for a team is split into eight periods.
First, let's see if the general pattern for the abilities looks right.
A team's abilities across periods share a common mean, so two ability
parameters for the same team should be correlated across iterations,
while parameters for different teams shouldn't show the same coupling.
Below are the pairs plots for the seventh and eighth periods for the
29th team and the first and second periods for the 30th team.

```{r plot_a_flat_pairs, dependson = "fit", out.width = "95%", fig.width = 8.14}
samps_flat <- as.array(fit)

prev_theme <- theme_update(panel.grid = element_blank(),
                           axis.text.y = element_blank(),
                           axis.text.x = element_blank())
color_scheme_set("gray")
mcmc_pairs(samps_flat,
           pars = c("a[7,29]", "a[8,29]", "a[1,30]", "a[2,30]"),
           off_diag_fun = "hex")
theme_set(prev_theme)
```

And here are the ability estimates for each team and period.

```{r info, cache.extra = file.info("../models/scorediff-split_2011.info.R")}
source("../lib/utils.R")
info <- source_as_list("../models/scorediff-split_2011.info.R")
```

```{r avals_plot, dependson = c("samps", "info"), fig.asp = 1}
n_periods <- 8
n_teams <- 30

sum_fn <- function(x){
    probs <- c(0.1, 0.25, 0.75, 0.9)
    qs <- quantile(x, probs = probs)
    names(qs) <- paste0("p", probs * 100)
    c(mean = mean(x), qs)
}

avals <- apply(samps$a, c(2, 3), sum_fn) %>%
    reshape2::melt(varnames = c("meas", "period", "team_idx")) %>%
    as_tibble() %>%
    mutate(team = info$team_names[team_idx]) %>%
    spread(meas, value)

teams_ranked <- arrange(info$wins_2011, desc(n_wins))$team

mutate(avals, team = factor(team, levels = teams_ranked)) %>%
    ggplot(aes(x = period, y = mean)) +
    geom_hline(yintercept = 0, color = tc$primary_light_dim) +
    geom_linerange(aes(ymin = p10, ymax = p90), size = 0.3) +
    geom_point() +
    facet_wrap(~ team) +
    labs(y = "mean ability (with 80% intervals)",
         title = "Team ability estimates for 2011 season",
         subtitle = "sorted by regular season standings") +
    theme(panel.grid.minor.y = element_blank(),
          panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank(),
          axis.text.x = element_blank())
```

It seems like the periods are capturing some variation, though many
teams might be sufficiently summarized by a single period ability
estimate.

There are some interesting teams, such as the Brewers, that look out
of place based on their season standings.  I'm not sure what's going
on here.  Assuming it's not a coding error, I suppose it should come
down to some combination of the observed match-ups (e.g., weaker
division) and iter-team variation in run differentials (e.g., winning
a disproportionate number of close games).

## Posterior predictive check

```{r dat, cache.extra = file.info("../models/scorediff-split_2011.data.R")}
dat <- source_as_list("../models/scorediff-split_2011.data.R")
```

```{r ppc, dependson = "fit"}
matchups <- tribble(~home, ~away,
                    "ARI", "COL",
                    "ATL", "WAS",
                    "CIN", "HOU",
                    "CLE", "MIN",
                    "LAN", "SFN",
                    "NYA", "BOS",
                    "PHI", "NYN",
                    "TEX", "SEA") %>%
    mutate(home_idx = match(home, info$team_names),
           away_idx = match(away, info$team_names))

observed_idx <- which(paste0(info$team_names[dat$team_home],
                             info$team_names[dat$team_away])
                      %in%
                      paste0(matchups$home, matchups$away))

observed <- tibble(score_diff = dat$score_diff[observed_idx],
                   home_idx = dat$team_home[observed_idx],
                   home_name = info$team_names[home_idx],
                   away_idx = dat$team_away[observed_idx],
                   away_name = info$team_names[away_idx],
                   home_period = factor(dat$period_home[observed_idx],
                                        levels = 1:8),
                   matchup = factor(paste0(away_name, "@", home_name)))

q95 <- function(x){
    qs <- quantile(x, probs = c(0.025, 0.975))
    names(qs) <- c("lower", "upper")
    qs
}

ppc <- apply(samps$score_diff_new[,observed_idx], 2, q95) %>%
    reshape2::melt(varnames = c("meas", "game")) %>%
    spread(meas, value) %>%
    mutate(home_name = observed$home_name,
           away_name = observed$away_name,
           home_period = factor(observed$home_period,
                                levels = 1:8),
           matchup = factor(paste0(away_name, "@", home_name))) %>%
    as_tibble() %>%
    group_by(matchup, home_period) %>%
    filter(row_number() == 1)
```

```{r ppc_plot, dependson = "ppc", out.width = "50%", fig.width = 4.28, fig.asp = 1.2}
ggplot(ppc, aes(x = home_period)) +
    geom_hline(yintercept = 0, color = tc$primary_light_dim) +
    geom_linerange(aes(ymin = lower, ymax = upper),
                   alpha = 0.8) +
    geom_point(aes(y = score_diff),
               fill = NA, color = tc$background_dark, shape = 21,
               data = observed) +
    scale_y_continuous(limits = c(-11, 11),
                       minor_breaks = -11:11) +
    labs(x = NULL, y = NULL,
         title = "Run differential",
         subtitle = "compared to the match-up's 95% predictive interval") +
    coord_flip() +
    facet_wrap(~ matchup, ncol = 1, scales = "free_y",
               strip.position = "left") +
    theme(panel.grid.major.y = element_blank(),
          panel.grid.minor.y = element_blank(),
          strip.text.y = element_text(angle = 180),
          axis.text.y = element_blank())
```

The interavls are still unimpressively large.  My next attempt to
shrink them will be to incorporate information about the starting
pitcher for each team.


------------------------------------------------------------------------

```{r session_info, echo = FALSE, results = "asis"}
source("session-info.R")
```
